---
title: "01_data_import"
author: "Ian Douglas"
date: "11/2/2019"
output: html_document
---
## Load required packages
```{r}
require(tidyverse)
require(haven)
```

## read raw SPSS data
```{r}
## ARCHIVED CHUNK (SKIP)
# age 3:
#D_wave3 = read_sav("../../data/raw/wave3/FF_wave3_2019.sav")
# age 9:
#D_wave5 = read_sav("../../data/raw/wave5/FF_wave5_2019.sav")
# age 15:
#D_wave6 = read_sav("../../data/raw/wave6/FF_wave6_2019.sav")

# save each as an .rds (faster for reading in next time)

# saveRDS(D_wave3,"../../data/raw/wave3/year3_Data.rds")
# saveRDS(D_wave5,"../../data/raw/wave5/year9_Data.rds")
# saveRDS(D_wave6,"../../data/raw/wave6/year15_Data.rds")
```

```{r}
D_wave3 = readRDS("../../data/raw/wave3/year3_Data.rds")
D_wave5 = readRDS("../../data/raw/wave5/year9_Data.rds")
D_wave6 = readRDS("../../data/raw/wave6/year15_Data.rds")
```

# Extract Age 3 'attachment style' categorical variable
```{r}
# store a dataframe containing the id number (for merging) and the variable of choice
attCat_df = D_wave3 %>%
  select(idnum, ch3att_codeabc)

# descriptives:
table(attCat_df$ch3att_codeabc)
#   -9    1    2    3 
# 2630   47 1719  502 

#dummy code the variable according to the variable documentation
attCat_df.1 = attCat_df %>%
  # first the three dummy codes:
  mutate(avoidant = ifelse(ch3att_codeabc == 1, 1, 0),
         secure = ifelse(ch3att_codeabc == 2, 1, 0),
         resistant = ifelse(ch3att_codeabc == 3, 1, 0)) %>%
  # now replace the numeric codes with a single column of labels (for grouping purposes)
  rowwise() %>%
  mutate(ch3att_codeabc = ifelse(all(c(avoidant, secure, resistant)==0), NA,
           c("avoidant", "secure", "resistant")[which(c(avoidant, secure, resistant)==1)])
  ) %>% ungroup() %>% # ungroup by each row
  # now convert `ch3att_codeabc` to a factor
  mutate_at(vars(ch3att_codeabc), ~factor(., levels = c("avoidant", "secure", "resistant")))
rm(list= c("D_wave3","attCat_df"))
```

# Extract the social anxiety and isolation variables from the other two waves
```{r}
# filter for the desired variables
### Note, column position 1 is the 'idnum' variable.
wave5 = D_wave5[c(1, #year 9 idnum
                  # equivalent to grep("^k5e1(a|b|c|d)$"):
                  grep("^k5e1([a-d])$", names(D_wave5)), #connectedness at school.
                  # SSRS:
                  grep("^t5b1([a-y])$", names(D_wave5)), # more SSRS:
                  grep("^t5b3([a-l])$", names(D_wave5)))]
# filter similarly at wave 6
wave6 = D_wave6[c(1, # year 15 idnum
                  grep("^k6b1([a-d])$", names(D_wave6)), # connectedness at school.
                  # SSRS:
                  grep("k6d1([a-l])", names(D_wave6)))]
```

## read in the meta data
```{r}
# # metadata
# metaData = read.csv("../../documentation/FFMetadata_v03.csv",
#                     stringsAsFactors = FALSE)
# # filter only variables at each wave:
# ## Wave 3
# saveRDS(filter(metaData, new_name %in% c("idnum","ch3att_codeabc")),
#         "../../documentation/wave3_year3_metaData.rds")
# ## Wave 5 (year 9)
# saveRDS(filter(metaData, new_name %in% names(wave5)),
#         "../../documentation/wave5_year9_metaData.rds")
# ## Wave 6 (year 15)
# saveRDS(filter(metaData, new_name %in% names(wave6)),
#         "../../documentation/wave6_year15_metaData.rds")
#rm("metaData")
# read these saved RDS files back in:
md_year3 = readRDS("../../documentation/wave3_year3_metaData.rds")
md_year9 = readRDS("../../documentation/wave5_year9_metaData.rds")
md_year15 = readRDS("../../documentation/wave6_year15_metaData.rds")
# select only columns with desired info.
md_year3_tbl = md_year3[c("new_name","wave","varlab","scale","respondent", "probe","qtext")]
md_year9_tbl = md_year9[c("new_name","wave","varlab","scale","respondent", "probe","qtext")]
md_year15_tbl =md_year15[c("new_name","wave","varlab","scale","respondent", "probe","qtext")]
#rm(list=c("md_year3","md_year9","md_year15"))
```

# dealing with NA
```{r}
## First, if a participant does not have attachment style data, they are dropped
keepID = attCat_df.1$idnum[!is.na(attCat_df.1$ch3att_codeabc)]
yr3_df = attCat_df.1[attCat_df.1$idnum %in% keepID, ]
yr9_df = wave5[wave5$idnum %in% keepID, ]
yr15_df = wave6[wave6$idnum %in% keepID, ]

# Next, at year 9
## Rename connectedness at school vars:
# :: coded 0-Not once ... 4-every day ::
names(yr9_df[2:5]) <- c("yr9_CNCT_PartOf", "yr9_CNCT_Close", "yr9_CNCT_Happy", "yr9_CNCT_Safe")
lapply(yr9_df[2:5], table) # just the connectdedness at school variables.
```

## Convert all less than 0 to NA, then create a sum score
### Only return NA if all are NA
```{r}
# Coding: 1 = not once ... 4 = everyday [less to more]
# Wording: Positive

# No reversal means higher score = MORE CONNECTED
Sum = function(x) {
  len = length(x)
  x2 = replace(x, x < 0, NA)
  out = ifelse(sum(is.na(x2)) == len, NA, mean(x2, na.rm = TRUE)*len)
  return(out)
}

yr9_df$yr9_CNCT_Sum <- as.numeric(apply(yr9_df[2:5], 1, Sum))
table(yr9_df$yr9_CNCT_Sum, useNA = "ifany")
```

## repeat for year 15
```{r}
## Rename connectedness at school vars (NOTE THE ORDER CHANGES for the first 2 items):
# :: coded 1-Strong Agree ... 4- Strong Disagree :: !!! 7 is 'homeschooled' !!!
names(yr15_df[2:5]) <-c("yr15_CNCT_CloseTo","yr15_CNCT_PartOf","yr15_CNCT_Happy","yr15_CNCT_Safe")
lapply(yr15_df[2:5], table) # just the connectdedness at school variables.
#table(yr15_df$yr15_CNCT_Sum, useNA = "ifany")
```

# remove if any kids were homeschooled at wave 15 (indicated by '-7').
### This is because they necessarily cannot report the Outcome variables (connectedness at sch.)
```{r}
# first some filtering based on the responses above
keepID = yr15_df$idnum[!as.logical(apply(yr15_df[2:5],1, function(x) {
  any(x==7)
}))]

# REPLACE the old data frames
yr3_df = yr3_df[yr3_df$idnum %in% keepID, ]
yr9_df = yr9_df[yr9_df$idnum %in% keepID, ]
yr15_df = yr15_df[yr15_df$idnum %in% keepID, ]
```

# coding connectedness at age 15
### REVERSE IT SO HIGHER SCORE IS MORE CONNECTED
```{r}
# As indicated above:
## coding: 1 = strong agree ... 4 = strong disagree
## wording: positive wording
## Conclusion: REVERSE CODE so higher number indicates more connection.
Sum_yr15 = function(x) {
  len = length(x)
  x2 = 5 - replace(x, x < 0, NA)
  out = ifelse(sum(is.na(x2)) == len, NA, mean(x2, na.rm = TRUE)*len)
  return(out)
}

# Now continue with the coding of CNCT
yr15_df$target = as.numeric(apply(yr15_df[2:5], 1, Sum_yr15))
table(yr15_df$target,useNA = "ifany")
```

# Compute the sum scores for the SSRS using the same algorithm
### The items need to be divided into subscales at year 9
### At year 15 a total sum score is the only metric obtainable
```{r}
# coding (for all three): 1=Never ... 4=Very often
# wording: social=POSITIVE; ext=NEGATIVE; int=NEGATIVE
# Conclusion: reverse the first one, not the other two.
# Note, Sum_yr15() is already set up to reverse a scale of max 4.
socialSkills = paste0("t5b1", letters[grep("[a-v]", letters)])
externalizing = paste0("t5b3", c("a","c","g","h","i","j"))
internalizing = paste0("t5b3", c("b","d","e","f","k","l"))
## start by replacing all negative numbers with NA:
yr9_df[c(socialSkills, externalizing, internalizing)] =
  lapply(yr9_df[c(socialSkills, externalizing, internalizing)], function(x) {
    replace(x, x < 0, NA)
  })
```

## Now compute the sum in the same way for the other form
#### Only reverse code "social skills" so a higher number is more dysfunction
```{r}
ssIndex = match(socialSkills, names(yr9_df))
extIndex = match(externalizing, names(yr9_df))
intIndex = match(internalizing, names(yr9_df))
# for social skills, it is on a 4 point scale, so it can re-use Sum_yr15()
# also, they are POSITIVELY worded, (but higher number is more agreement)
# so the Sum_yr15 is perfect as it will also reverse code.
yr9_df$rev_SocSkillSum <- 
  as.numeric(apply(yr9_df[ssIndex], 1, Sum_yr15))
# For the other two scales, no reversal is needed, so use Sum()
yr9_df$externalizingSum <- 
  as.numeric(apply(yr9_df[extIndex], 1, Sum))
yr9_df$internalizingSum <- 
  as.numeric(apply(yr9_df[intIndex], 1, Sum))

# Finally, code a master sum across the three categories
yr9_df$socialSumTOTAL <- as.numeric(
  apply(yr9_df[(ncol(yr9_df)-2):ncol(yr9_df)], 1, Sum
))
```

## Compute the sum at year 15 for sociality
### To keep this consistent with year 9 (which includes internal- and externalizing symptoms), sociality score will be higher if a subject is more antisocial.
```{r}
# coding: 1=not true ... 3=Often true
## All the items are positively worded at year 15, so they will be reversed
## since the coding is also (disagree ... agree)
## They are also on a 3 point scale, so a new function is needed:
sum_rev3 = function(x) {
  len = length(x)
  x2 = 4 - replace(x, x < 0, NA)
  out = ifelse(sum(is.na(x2)) == len, NA, mean(x2, na.rm = TRUE)*len)
  return(out)
}
NCOL = ncol(yr15_df) # recall the last column is now 'target'
yr15_df$rev_socSkill <- as.numeric(apply(yr15_df[-c(1:5,NCOL)], 1, sum_rev3))
```

# Merge data
```{r}
newData = left_join(yr3_df, 
                    select(yr9_df, idnum, yr9_CNCT_Sum:socialSumTOTAL),
                    by = "idnum") %>%
  left_join(., select(yr15_df, idnum, rev_socSkill, target),by = "idnum")
```

Summary: The outcome (and its level at age 9) are positively coded, while the predictors are the opposite. Connectedness at school is reflected in a higher connectedness score at both timepoints. For sociality (social skills), a higher score is more 'antisociality' (roughly defined), and for internalizing/externalizing a higher score is more problem internal- and externalization.
```{r}
saveRDS(newData, "../../data/processed/modelData.rds")
```


