---
title: "02_analysis"
author: "Ian Douglas"
date: "12/12/2019"
output: html_document
---
### Analysis Plan
1. Compute correlations between the outcome, age 15 connectedness at school (CS15), and the predictors: internalizing behaviors (IB) and age 9 (CS9).
2. Examine mean differences between the groups, and variance ratios.
3. Due to the "rule of thumb", conclude that mean differences do not differ, but variance homogeneity is likely violated.
4. Conduct Welch's test to control Type I error rate while assessing the effect of the grouping variable.
5. The mean differences were found to be non-significant, but the hypothesis purports a crossing-type interaction, so plot the groups on the dimensions of IB (binned into high and low categories) and CS15.
6. The graphical evidence is clear that an interaction exists, and that a meaningful grouping of subjects should be 'avoidant' and 'not avoidant'. 
7. Fit the interaction model: predict CS15 from the Group-by-IB interaction, controlling for the continuous covariate CS9.
8. Report the F-statistic and p-value to indicate the significance of the model, as well as the significant (and standardized) Beta coefficients as a measure of effect size for the significant covariate and the interaction.
9. Interpret the conclusion of the model: the significant interaction shows that after controlling for CS9, the negative effect of IB on CS15 is worsened for subjects exhibiting avoidant attachment with caregivers at age three.

### Load libraries
```{r}
require(effsize)
require(tidyverse)
require(car)
```

### Read in data
```{r}
# convert from Tibble to data.frame
df = as.data.frame(readRDS("../../data/processed/modelData.rds"))
rownames(df) <- df$idnum
```

# Compute the correlations between IB and CS9 with CS15.
```{r}
# the continuous predictor
cor.test(df$target, df$internalizingSum, use = "complete.obs")
# the year9 observation of outcome (to control for)
cor.test(df$target, df$yr9_CNCT_Sum, use = "complete.obs")
# testing multicolinearity
cor.test(df$internalizingSum, df$yr9_CNCT_Sum, use = "complete.obs")
```
 Both are significantly correlated (though weakly) with the outcome. Weak multicolinearity present as well.

# Standardized mean differences and variance ratios.
## Pairwise comparison between the three group individually on the outcome (CS15)
```{r}
secure_avoidant = df %>%
  filter(resistant == 0)
resistant_secure = df %>%
  filter(avoidant == 0)
resist_avoid = df %>%
  filter(secure == 0)
contrast_list = list(secure_avoidant, resistant_secure, resist_avoid)
effsize::cohen.d(d = contrast_list[[1]]$target, as.factor(contrast_list[[1]]$secure),
                 na.rm = TRUE)
effsize::cohen.d(d = contrast_list[[2]]$target, as.factor(contrast_list[[2]]$secure),
                 na.rm = TRUE)
effsize::cohen.d(d = contrast_list[[3]]$target, as.factor(contrast_list[[3]]$avoidant),
                 na.rm = TRUE)
```

## Pairwise comparison between the three group individually on covariate IB
```{r}
# secure v avoidant
effsize::cohen.d(d = contrast_list[[1]]$internalizingSum,
                 as.factor(contrast_list[[1]]$secure),
                 na.rm = TRUE)
# secure v resistant
effsize::cohen.d(d = contrast_list[[2]]$internalizingSum,
                 as.factor(contrast_list[[2]]$secure),
                 na.rm = TRUE)
# avoidant v resistant
effsize::cohen.d(d = contrast_list[[3]]$internalizingSum,
                 as.factor(contrast_list[[3]]$avoidant),
                 na.rm = TRUE)
```
None of the means differ on average between groups.

## Variance ratios for CS15, CS9, and IB
```{r}
# Avoidant groups relative to not, for age 15 connectedness
var(na.omit((filter(df, avoidant == 1))$target))/var(na.omit((filter(df, avoidant == 0))$target))
# Age 9 connectedness
var(na.omit((filter(df, avoidant == 1))$yr9_CNCT_Sum))/var(na.omit((filter(df, avoidant == 0))$yr9_CNCT_Sum))
# IB
var(na.omit((filter(df, avoidant == 1))$internalizingSum))/var(na.omit((filter(df, avoidant == 0))$internalizingSum))
```

Results indicate a degree of imbalance that is concerning, according to Rubin's (2001) guidelines, for CS measured at age 9 and 15 (but not the covariate IB). The scenario suggests that the group with a much smaller sample size, those of avoidant attachment styles, has the much higher variance in the domain of school connectedness at ages 9 and 15 than the other two groups. Thus, conducting traditional ANOVA is more likely to increase the risk of comitting Type I error. Instead, to test the significance of the grouping factor on CS15, I will use Welch's test.

## Visualization
```{r}
# A density plot wil visualize the means and variances for each group as well:
#pdf(file="../../plots/groupDensities.pdf", width = 12, height = 8)
densityPlot(df$target, g = df$ch3att_codeabc, 
            legend = list(location = "topleft", title = 'Attachment Style'),
            xlab = "Connectedness at School", adjust = c(.7, 2, 2))
abline(v=mean(na.rm=TRUE,(df%>%filter(ch3att_codeabc=="avoidant"))$target), lwd =2)
abline(v=mean(na.rm=TRUE,(df%>%filter(ch3att_codeabc=="secure"))$target),lty=2,col=4,lwd =2)
abline(v=mean(na.rm=TRUE,(df%>%filter(ch3att_codeabc=="resistant"))$target),lty=3,col=6,lwd =2)
#dev.off()

## And with two groups
density_Data = df %>% mutate_at(vars(avoidant), ~as.factor(ifelse(.==1,"avoidant","other")))
#pdf(file="../../plots/binTWO_groupDensities.pdf", width = 12, height = 8)
densityPlot(density_Data$target, g = density_Data$avoidant,
            legend = list(location = "topleft", title = 'Avoidant'),
            xlab = "Connectedness at School", adjust = c(.7, 2.3))
abline(v = mean(na.rm=TRUE,(density_Data%>%filter(avoidant=="avoidant"))$target), lwd =2)
abline(v=mean(na.rm=TRUE,(density_Data%>%filter(avoidant=="other"))$target),lty=2,col=4,lwd =2)
#dev.off()
```

# Welch's one-way ANOVA corection for heteroskedasticity.
## Comparing the avoidant to the resistant group
```{r}
oneway.test(target ~ as.factor(avoidant), data = resist_avoid)
```

## Comparing the avoidant to the secure group
```{r}
oneway.test(target ~ as.factor(avoidant), data = secure_avoidant)
```

## Comparing the avoidant to the non-avoidant group
```{r}
oneway.test(target ~ as.factor(avoidant), data = df)
```

## All groups (omnibus):
```{r}
oneway.test(target ~ ch3att_codeabc, data = df)
```

Conclusion: the groups alone do not explain a significant proportion in the variance of CS15.

# Plot the hypothesized interaction.
## At low and high levels of internalizing, plot group means for target
```{r}
plt_data = df %>%
  filter(!is.na(internalizingSum)) %>%
  group_by(ch3att_codeabc) %>%
  mutate(internalizing = ifelse(internalizingSum <= median(internalizingSum,na.rm = TRUE),
                                  "low","high")) %>%
  group_by(ch3att_codeabc, internalizing) %>%
  summarize_at(vars(internalizingSum, target), 
               list(~mean(., na.rm = TRUE), ~sd(., na.rm = TRUE))) %>%
  ungroup() %>%
  rename_at(vars(ch3att_codeabc), ~("Group"))
plt <- ggplot(plt_data, aes(x = internalizing, y = target_mean, group = Group)) +
  geom_point(aes(color = Group)) +
  geom_line(aes(color = Group)) +
  ggtitle(label = "Mean Connectedness at School by Group and Level of Internalizing Behavior")
# ggsave(plt, device = "pdf", file="../../plots/interaction_plot.pdf",
#        width = 12, height = 8)
plt
```


# binning to pit avoidant against all other groups, as per theory
```{r}
plt_data2 = df %>% 
  mutate_at("avoidant", ~as.factor(ifelse(. == 1, "True", "False"))) %>%
  filter(!is.na(internalizingSum)) %>%
  mutate(internalizing = ifelse(internalizingSum <= median(internalizingSum,na.rm = TRUE),
                                  "low","high")) %>%
  group_by(avoidant, internalizing) %>%
  summarize_at(vars(internalizingSum, target), 
               list(~mean(., na.rm = TRUE), ~sd(., na.rm = TRUE))) %>%
  ungroup()
plt2 <- ggplot(plt_data2, aes(x = internalizing, y = target_mean, group = avoidant)) +
  geom_point(aes(color = avoidant)) +
  geom_line(aes(color = avoidant)) +
  ggtitle(label = "Mean Connectedness at School by Group and Level of Internalizing Behavior")
# ggsave(plt, device = "pdf", file="../../plots/2group_int_plot.pdf",
#        width = 12, height = 8)
plt2
```

This crossing type interaction is a possible explanation for the undeteceted mean differences, and insignificant main effect of group (both when binning created two- and three-groups). This provides graphical evidence that an interaction model may still be appropriate, and subsequent modeling (below) will follow from this conclusion.

# Regression Analyses
## Fit the model
```{r}
model = lm(target ~ yr9_CNCT_Sum + # year 15 connectedness controlling for that of year 9
             internalizingSum*avoidant, # predicted by the int. btwn internalizing and avoidant
           data = df)
```

## Regression estimates
```{r}
summary(model)
```

## ANOVA table:
```{r}
summary(aov(model))
```


